---
title: "unused slides"
format: revealjs
---

# unused 

## Imputation

```{r}
# dat[is.na(dat)] <- 0.5
# ggmice(dat, aes(x,y)) +
#   geom_point(size = 3) +
#   labs(x = "", y = "")
```

## Example data

```{r, echo=FALSE}
dat_raw <- read.csv("~/Downloads/table_2CE4F5AC7A7A48E9A5AFC8D7E2DF7C9C.csv", header=FALSE)
nms <- as.character(dat_raw[1, -1])
dat_pre <- dat_raw[-1, -1] |> setNames(nms)
dat_pre[dat_pre == ""] <- NA
dat_wide <- dat_pre |> 
  na.omit() |>
  t() |>
  as.data.frame()

row.names(dat_wide)[-1] 
age_cats <- c("15-24", "25-44", "45-64", "65+")
names(age_cats) <- age_cats
queer_cats <- c(rep(1, length(age_cats)), rep(0, length(age_cats))) 

dat <- data.frame(age = rep(age_cats, 2), LGBTQIA = factor(queer_cats, levels = 0:1, labels = c("no", "yes")), prop = c(dat_wide[-1, 1], dat_wide[-1, 2]))  

ggplot(dat, aes(age, prop, color = LGBTQIA)) +
  geom_point() +
  ggmice:::theme_mice()

```


## Introduction

This is a Quarto presentation format that is consistent with the rainbowR [brand](https://rainbowr.org/new-look).


:::{.notes}
Speaker notes go here.
:::

# inverse class for section headings {.inverse}

## {.inverse .center-h .center .larger125}

inverse class is also useful for stressing [*key points*]{.green}. 

## Colours

There's a colour for inline code such as `x <- c(1, 3, 6)` and `{.inverse}`.

There's also a colour for short [links](https://github.com/r-lgbtq/rainbowrpres) and full ones like <https://quarto.org/docs/presentations/revealjs/>.

## Colours {.inverse}

There's a colour for inline code such as `x <- c(1, 3, 6)` and `{.inverse}`.

There's also a colour for short [links](https://github.com/r-lgbtq/rainbowrpres) and full ones like <https://quarto.org/docs/presentations/revealjs/>.

- The inline code and link colours change when the inverse class is applied.

## Colour classes for text {.inverse .larger150}

[.red]{.red}

[.orange]{.orange}

[.yellow]{.yellow}

[.green]{.green}

[.blue]{.blue}

[.pink]{.pink}

## Colour classes for highlighting {.larger150}

[.red-hl]{.red-hl}

[.orange-hl]{.orange-hl}

[.yellow-hl]{.yellow-hl}

[.green-hl]{.green-hl}

[.blue-hl]{.blue-hl}

[.pink-hl]{.pink-hl}

## Yellow background {.yellow-bg footer=false}

`.yellow-bg` class, and similar for the other rainbowR colours.

## Code

The extension includes a custom syntax highlighting theme:

```{r}
#| echo: true
#| eval: false
library(dplyr)
library(palmerpenguins)

penguins %>%
  mutate(
    long_flipper = case_when(
      species == "Adelie" & flipper_length_mm > 195 ~ TRUE,
      species == "Chinstrap" & flipper_length_mm > 200 ~ TRUE,
      species == "Gentoo" & flipper_length_mm > 225 ~ TRUE,
      TRUE ~ FALSE
    )
  ) %>%
  mutate(
    body_mass_kg = body_mass_g / 1000,
    bill_length_cm = bill_length_mm / 10,
    bill_depth_cm = bill_depth_mm / 10
  )
```

## Callouts {.smaller80}

There are custom colours for callouts: 

:::{.callout-note}
A note
:::

:::{.callout-tip}
This is a good idea
:::

:::{.callout-important}
Something important
:::

:::{.callout-caution}
Be careful about doing this
:::

:::{.callout-warning}
A warning
:::

## Callouts {.smaller80 .inverse}

There are custom colours for callouts: 

:::{.callout-note}
A note
:::

:::{.callout-tip}
This is a good idea
:::

:::{.callout-important}
Something important
:::

:::{.callout-caution}
Be careful about doing this
:::

:::{.callout-warning}
A warning
:::

## Columns with backgrounds

::: {columns}
::: {.column .col-box-yellow width="32%"}
### Box 1

Use class `.col-box-{color}` to apply colour box styling, e.g. `.col-box-yellow`
:::

::: {.column .col-box-orange width="32%"}
### Box 2

Default height is 450px
:::

::: {.column .col-box-green width="32%"}
### Box 3
:::
:::

## Columns with backgrounds: tall and short

::::: {columns}
:::: {.column .col-box-yellow-tall width="49%"}
Colour box styling can be applied to entire columns

This column has class `.col-box-yellow-tall`
::::

:::: {.column width="49%"}
:::{.col-box-orange-short}
Or to divs within columns to create stacks
:::

:::{.col-box-green-short}
This div has class `.col-box-green-short`
:::

::::

:::::

## Columns with backgrounds: tall and short {.inverse}

::::: {columns}
:::: {.column .col-box-red-tall width="49%"}
Colour box styling can be applied to entire columns

This column has class `.col-box-red-tall`
::::

:::: {.column width="49%"}
:::{.col-box-blue-short}
Or to divs within columns to create stacks
:::

:::{.col-box-pink-short}
This div has class `.col-box-pink-short`
:::

::::

:::::

# Non-rainbowR-specific nice-to-haves {.inverse}

## Alignment

There are classes to help with text alignment:

:::{.center-h}
`{.center-h}` gives horizontally-centered text
:::

:::{.right}
`{.right}` gives right-aligned text
:::

:::{.callout-note}
For all revealjs Quarto presentations, applying `{.center}` to the slide title vertically centers the whole slide.
:::

## Sizing

There are classes to help size text by a given percentage:

[`{.smaller90}`]{.smaller90}, [`{.smaller80}`]{.smaller80}, [`{.smaller75}`]{.smaller75}
[`{.smaller70}`]{.smaller70}, [`{.smaller60}`]{.smaller60}, [`{.smaller50}`]{.smaller50}

[`{.larger125}`]{.larger125}, [`{.larger150}`]{.larger150}, [`{.larger175}`]{.larger175}, 
[`{.larger200}`]{.larger200}, [`{.larger250}`]{.larger250}

## More information

You can learn more about Quarto Revealjs presentations at <https://quarto.org/docs/presentations/revealjs/>

You can learn more about Quarto extensions at <https://quarto.org/docs/extensions/>

You can find these slides at <https://r-lgbtq.github.io/rainbowrpres/>
